---
title: "Mark's Spawn Timing"
author: "Sarah Gaichas"
date: "2024-01-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Mark Wuenschel provided the data and code to produce the figures from working papers on spawning phenology of haddock and yellowtail flounder stocks.

Mark's input data and code are in this repo.

I just ran Mark's code and will save them as rds files for input into ecodata.

## Yellowtail

Input data: [`YT_63_23_BTS.csv`]()

Processing code: [`YT_Prop_Mat_by_temp_week_timeblocks_CC_WP.R`]()

I ran the code from lines 1:219 to get the datasets. 

```{r}
## program to read data files,
# plot proportions pre -spawn-post spawn to evaluate changes in spawning phenology during SBTS
####condense proportions maturity by temp and Day of year
# last modified 10/10/2018

rm(list=ls(all=TRUE))  # clear the workspace of all existing variables
#setwd("C:\\R\\Spawning_Phenology") # set the path for the working directory
#od= 'C:/R/Spawning_Phenology/output/'   #directory for writing output

library(MASS)

YTdata=as.data.frame(read.csv(file="YT_63_23_BTS.csv", header=T, sep=","))
#Allspp_SPdata$mean[which(Allspp_SPdata$mean==0)] <- NA
data=subset(YTdata[YTdata$SVSPP=='105',]) #select species

#maxFL<-max(as.numeric(data[,7])) #need to change to appropriate column in new dataset

###add in stock by specifying strata
GBSTRAT<- c('1130','1140','1150','1160','1170','1180','1190','1200','1210')
SNESTRAT_SP<- c('1010','1020','1050','1060','1090','1100','1690','1730','1740')
SNESTRAT_FA<- c('1010','1020','1050','1060','1090','1100')
CCSTRAT_SP<- c('1250','1260','1270','1390','1400','3560','3570','3590','3600','3610','3620','3640','3650','3660')
CCSTRAT_FA<- c('1250','1260',       '1390','1400','3560','3570','3590','3600','3610','3620','3640','3650','3660')


data$STOCK <-NA

sel <-which(data$STRATUM %in% GBSTRAT)
data$STOCK[sel]<- with (data[sel,], 'GB')

sel <-which(data$STRATUM %in% SNESTRAT_SP)
data$STOCK[sel]<- with (data[sel,], 'SNE')


sel <-which(data$STRATUM %in% CCSTRAT_SP)
data$STOCK[sel]<- with (data[sel,], 'CC')

###  summary(as.factor(data$STOCK))

###add in day of year from date
data$DOY <-strftime(as.Date(data$BEGIN_EST_TOWDATE, "%d-%b-%y"), format = "%j")
data$DOY<- as.numeric(data$DOY)


### calculate proportions pre-spawning, spawning active, post-spawning in each season/year/stock
###drop males and immatures and unknowns
data=subset(data[data$EST_YEAR!='1977',])
data=subset(data[data$SEX=='2',])
data=subset(data[data$MATURITY!='Unknown',])
data=subset(data[data$MATURITY!='Immature',])
data=subset(data[data$EST_YEAR!=1969,])



data$SPAWNCOND <- NA
sel <-which(data$MATURITY=="Developing" )
data$SPAWNCOND[sel]<- with (data[sel,], 'Prespawning')
sel <-which(data$MATURITY=="Ripe"|data$MATURITY=="Ripe and Running" )
data$SPAWNCOND[sel]<- with (data[sel,], 'Spawning')
sel <-which(data$MATURITY=="Spent"|data$MATURITY=="Resting" )
data$SPAWNCOND[sel]<- with (data[sel,], 'Postspawning')

sel <-which(data$MATURITY=="Ripe"|data$MATURITY=="Ripe and Running" )
data$MATURITY[sel]<- with (data[sel,], 'Ripe')

data$EST_YEAR <- as.factor(data$EST_YEAR)
data$STOCK <- as.factor(data$STOCK)
data$PROPSPAWN <- NA
data$TEMPBIN <-round(data$BOTTEMP, digits=0)
data$TEMPBIN <-as.factor(data$TEMPBIN)
data$WOY <- data$DOY/7
data$WOY <-round(data$WOY, digits=0)
data$WOY <-as.factor(data$WOY)
data<-(data[!is.na(data$TEMPBIN),])

##concatenate cruise 6 and stock to ease calculations of proportions
#data$CRUISE6_STOCK <- paste(data$CRUISE6,"_",data$STOCK)

data$YEARBLOCK <- NA
YB70s<- c('1970','1971','1972','1973','1974','1975','1976','1977','1978','1979')
YB80s<- c('1980','1981','1982','1983','1984','1985','1986','1987','1988','1989')
YB90s<- c('1990','1991','1992','1993','1994','1995','1996','1997','1998','1999')
YB00s<- c('2000','2001','2002','2003','2004','2005','2006','2007','2008','2009')
YB10s<- c('2010','2011','2012','2013','2014','2015','2016','2017','2018','2019')
YB20s<- c('2020','2021','2022','2023','2024','2025','2026','2027','2028','2029')
sel <-which(as.character(data$EST_YEAR) %in% YB70s )
data$YEARBLOCK[sel]<- with (data[sel,], 'YB70s')
sel <-which(as.character(data$EST_YEAR) %in% YB80s )
data$YEARBLOCK[sel]<- with (data[sel,], 'YB80s')
sel <-which(data$EST_YEAR %in% YB90s )
data$YEARBLOCK[sel]<- with (data[sel,], 'YB90s')
sel <-which(data$EST_YEAR %in% YB00s )
data$YEARBLOCK[sel]<- with (data[sel,], 'YB00s')
sel <-which(data$EST_YEAR %in% YB10s )
data$YEARBLOCK[sel]<- with (data[sel,], 'YB10s')
sel <-which(data$EST_YEAR %in% YB20s )
data$YEARBLOCK[sel]<- with (data[sel,], 'YB20s')

#data2=subset(data[data$STOCK==c("CC","SNE","GB"),])
data2<- data[!is.na(data$STOCK), ]

#data$PROPSPAWN<- rowsum(data$SPAWNCOND, c(data$EST_YEAR,data$SEASON,dta$STOCK), na.rm=T)
###loop to calculateproportions for each year, season, stock


####cleaning up the data a little#####
###may need to remove some small fish that are listed as mature (<20 cm) seems unlikely. 
#library(plyr); 
library(dplyr)
# results.by.year.season.stock <- ddply(.data=data, .var=C(data$EST_YEAR,data$SEASON, data$STOCK), .fun=function(x) {
#   data.frame(n=nrow(x),
#              num_prespawn = nrow(subset(x, data$SPAWNCOND %in%
#                                           "Prespawning")) 
#              
#              #prop_prespawn= nrow(subset(x, data$SPAWNCOND %in%
#              #                             "Prespawning")) / nrow(x)
#             )
# })


#results.by.year.season.stock <- ddply(data, C(data$EST_YEAR,data$SEASON, data$STOCK), mutate, nprespawn=(sum(count(data$SPAWNCOND=="Prespawning"))))


###aggregate by year and season

####this parts gets the summary date and temp data
data3<- data2 %>%
  group_by (SEASON, STOCK, TEMPBIN, YEARBLOCK, MATURITY) %>%
  summarize(minJDAY= min(DOY),
            meanJDAY= mean(DOY),
            maxJDAY=max(DOY),
            meanLENGTH = mean(LENGTH),
            meanTEMP=mean(BOTTEMP, na.rm=T),
            #Prespawning= n (SPAWNCOND=="Prespawning"),
            #Prespawning= length (SPAWNCOND=="Prespawning"),
            totalMF=n()
  )


###this part tallies by spawn condition and adds colums for each, so there is one row per yearxseasonxstock
library(reshape2)
data3<-dcast(data3,SEASON+STOCK+TEMPBIN+YEARBLOCK~ MATURITY, value.var="totalMF")


data4<- data2 %>%
  group_by ( SEASON, STOCK, TEMPBIN,YEARBLOCK) %>%
  summarise(minJDAY= min(DOY),
            meanJDAY= mean(DOY),
            maxJDAY=max(DOY),
            meanLENGTH = mean(LENGTH),
            meanTEMP=mean(BOTTEMP, na.rm=T)
            #Prespawning= n (SPAWNCOND=="Prespawning"),
            #Prespawning= length (SPAWNCOND=="Prespawning"),
            #totalMF=n()
  )

### merging to tack on the bio and env data to the summary of spawning info
data5<- merge(data3, data4, by = c( "SEASON", "STOCK","TEMPBIN", "YEARBLOCK"))
### add in columns for proportions
data5$Developing[is.na(data5$Developing)]<-0
data5$Ripe[is.na(data5$Ripe)]<-0
data5$Spent[is.na(data5$Spent)]<-0
data5$Resting[is.na(data5$Resting)]<-0

data5$PROPD<- 100*data5$Developing/(data5$Developing+data5$Ripe+data5$Spent+data5$Resting)
data5$PROPR<- 100*data5$Ripe/(data5$Developing+data5$Ripe+data5$Spent+data5$Resting)
data5$PROPS<- 100*data5$Spent/(data5$Developing+data5$Ripe+data5$Spent+data5$Resting)
data5$PROPT<- 100*data5$Resting/(data5$Developing+data5$Ripe+data5$Spent+data5$Resting)
data5$MF<-(data5$Developing+data5$Ripe+data5$Spent+data5$Resting )
#data$mean[which(data$mean==0)] <- NA


#####doing similar merge to get WOY summary
####this parts gets the summary date and temp data
WOYdata3<- data2 %>%
  group_by (SEASON, STOCK, WOY, MATURITY, YEARBLOCK) %>%
  summarize(minJDAY= min(DOY),
            meanJDAY= mean(DOY),
            maxJDAY=max(DOY),
            meanLENGTH = mean(LENGTH),
            meanTEMP=mean(BOTTEMP, na.rm=T),
            #Prespawning= n (SPAWNCOND=="Prespawning"),
            #Prespawning= length (SPAWNCOND=="Prespawning"),
            totalMF=n()
  )


###this part tallies by spawn condition and adds colums for each, so there is one row per yearxseasonxstock
library(reshape2)
WOYdata3<-dcast(WOYdata3,SEASON+STOCK+WOY+YEARBLOCK~ MATURITY, value.var="totalMF")


WOYdata4<- data2 %>%
  group_by ( SEASON, STOCK, WOY,YEARBLOCK) %>%
  summarise(minJDAY= min(DOY),
            meanJDAY= mean(DOY),
            maxJDAY=max(DOY),
            meanLENGTH = mean(LENGTH),
            meanTEMP=mean(BOTTEMP, na.rm=T)
            #Prespawning= n (SPAWNCOND=="Prespawning"),
            #Prespawning= length (SPAWNCOND=="Prespawning"),
            #totalMF=n()
  )

### merging to tack on the bio and env data to the summary of spawning info
WOYdata5<- merge(WOYdata3, WOYdata4, by = c( "SEASON", "STOCK","WOY","YEARBLOCK"))


### add in columns for proportions
WOYdata5$Developing[is.na(WOYdata5$Developing)]<-0
WOYdata5$Ripe[is.na(WOYdata5$Ripe)]<-0
WOYdata5$Spent[is.na(WOYdata5$Spent)]<-0
WOYdata5$Resting[is.na(WOYdata5$Resting)]<-0

WOYdata5$PROPD<- 100*WOYdata5$Developing/(WOYdata5$Developing+WOYdata5$Ripe+WOYdata5$Spent+WOYdata5$Resting)
WOYdata5$PROPR<- 100*WOYdata5$Ripe/(WOYdata5$Developing+WOYdata5$Ripe+WOYdata5$Spent+WOYdata5$Resting)
WOYdata5$PROPS<- 100*WOYdata5$Spent/(WOYdata5$Developing+WOYdata5$Ripe+WOYdata5$Spent+WOYdata5$Resting)
WOYdata5$PROPT<- 100*WOYdata5$Resting/(WOYdata5$Developing+WOYdata5$Ripe+WOYdata5$Spent+WOYdata5$Resting)
WOYdata5$MF<-(WOYdata5$Developing+WOYdata5$Ripe+WOYdata5$Spent+WOYdata5$Resting )
#data$mean[which(data$mean==0)] <- NA
##break data into spring and fall separately for plotting and analysis

```

To get annual percent by spawning stage, I'll start with data2 and group by year instead of yearblock but otherwise use the same code:

```{r}

###aggregate by year and season

####this parts gets the summary date and temp data
data3yr<- data2 %>%
  group_by (SEASON, STOCK, EST_YEAR, MATURITY) %>%
  summarize(minJDAY= min(DOY),
            meanJDAY= mean(DOY),
            maxJDAY=max(DOY),
            meanLENGTH = mean(LENGTH),
            meanTEMP=mean(BOTTEMP, na.rm=T),
            #Prespawning= n (SPAWNCOND=="Prespawning"),
            #Prespawning= length (SPAWNCOND=="Prespawning"),
            totalMF=n()
  )

data3yr<-dcast(data3yr,SEASON+STOCK+EST_YEAR~ MATURITY, value.var="totalMF")

data4yr<- data2 %>%
  group_by ( SEASON, STOCK, EST_YEAR) %>%
  summarise(minJDAY= min(DOY),
            meanJDAY= mean(DOY),
            maxJDAY=max(DOY),
            meanLENGTH = mean(LENGTH),
            meanTEMP=mean(BOTTEMP, na.rm=T)
            #Prespawning= n (SPAWNCOND=="Prespawning"),
            #Prespawning= length (SPAWNCOND=="Prespawning"),
            #totalMF=n()
  )

data5yr<- merge(data3yr, data4yr, by = c( "SEASON", "STOCK", "EST_YEAR"))

### add in columns for proportions
data5yr$Developing[is.na(data5yr$Developing)]<-0
data5yr$Ripe[is.na(data5yr$Ripe)]<-0
data5yr$Spent[is.na(data5yr$Spent)]<-0
data5yr$Resting[is.na(data5yr$Resting)]<-0

data5yr$PROPD<- 100*data5yr$Developing/(data5yr$Developing+data5yr$Ripe+data5yr$Spent+data5yr$Resting)
data5yr$PROPR<- 100*data5yr$Ripe/(data5yr$Developing+data5yr$Ripe+data5yr$Spent+data5yr$Resting)
data5yr$PROPS<- 100*data5yr$Spent/(data5yr$Developing+data5yr$Ripe+data5yr$Spent+data5yr$Resting)
data5yr$PROPT<- 100*data5yr$Resting/(data5yr$Developing+data5yr$Ripe+data5yr$Spent+data5yr$Resting)
data5yr$MF<-(data5yr$Developing+data5yr$Ripe+data5yr$Spent+data5yr$Resting )


```

Plot annual percentage by stage

```{r}

annual <- data5yr |> 
  dplyr::filter(SEASON %in% c("SPRING", "FALL")) |>
  dplyr::mutate(Time = as.numeric(EST_YEAR)) |>
  dplyr::select(SEASON, STOCK, Time, 
                Developing = PROPD, 
                Ripe = PROPR, 
                Spent = PROPS, 
                Resting = PROPT) |>
  tidyr::pivot_longer(Developing:Resting, names_to = "Stage", values_to = "Value") 

mat.col=c("orange", "cyan", "purple", "coral3")
names(mat.col) <- as.factor(c("Developing", "Ripe", "Spent", "Resting"))



p <- ggplot2::ggplot(annual, 
                     ggplot2::aes(x=Time, y=Value, fill = as.factor(Stage))) + # , colour = Stage
  #ggplot2::geom_point() +
  #ggplot2::geom_line() +
  #ecodata::geom_gls() +
  ggplot2::geom_bar(stat="identity", position="fill") +
  ggplot2::scale_fill_manual(values = mat.col) +
  ggplot2::facet_wrap(SEASON~STOCK) + 
  ecodata::theme_facet() +
  ggplot2::ggtitle("Yellowtail flounder stocks")

p
```

Plot time block and temp bin percentage by stage

```{r}
decadal <- data5 |> 
  dplyr::filter(SEASON %in% c("SPRING", "FALL")) |>
  dplyr::mutate(Time = YEARBLOCK) |>
  dplyr::select(SEASON, STOCK, Time, PROPD, PROPR, PROPS, PROPT) |>
  tidyr::pivot_longer(PROPD:PROPT, names_to = "Stage", values_to = "Value")

p <- ggplot2::ggplot(decadal, # #|>dplyr::filter(Stage == "PROPT"), 
                     ggplot2::aes(x=Time, y=Value, colour = Stage)) + # , colour = Stage
  ggplot2::geom_point() +
  ggplot2::geom_line() +
  #ecodata::geom_gls() +
  ggplot2::facet_wrap(SEASON~STOCK) + 
  ecodata::theme_facet()

p
```

Plot temperatures

```{r}
annual <- data5yr |> 
  dplyr::filter(SEASON %in% c("SPRING", "FALL")) |>
  dplyr::mutate(Time = as.numeric(EST_YEAR)) |>
  dplyr::select(SEASON, STOCK, Time, meanTEMP) 

p <- ggplot2::ggplot(annual, 
                     ggplot2::aes(x=Time, y=meanTEMP)) + # , colour = Stage
  ggplot2::geom_point() +
  ggplot2::geom_line() +
  ecodata::geom_gls() +
  ggplot2::facet_wrap(SEASON~STOCK) + 
  ecodata::theme_facet() +
  ggplot2::ggtitle("Yellowtail flounder stocks")

p
```
Day of year

```{r}
annual <- data5yr |> 
  dplyr::filter(SEASON %in% c("SPRING", "FALL")) |>
  dplyr::mutate(Time = as.numeric(EST_YEAR)) |>
  dplyr::select(SEASON, STOCK, Time, meanJDAY) 

p <- ggplot2::ggplot(annual, 
                     ggplot2::aes(x=Time, y=meanJDAY)) + # , colour = Stage
  ggplot2::geom_point() +
  ggplot2::geom_line() +
  ecodata::geom_gls() +
  ggplot2::facet_wrap(SEASON~STOCK) + 
  ecodata::theme_facet() +
  ggplot2::ggtitle("Yellowtail flounder stocks")

p
```

